Parallel-in-time methods for Particle Simulations

### 2024-12-05 paper/main.tex
Paper: edits

\begin{abstract}
The purpose of this paper is to examine the applications of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration. We then test and measure the real speed-up and compare it with the expectation derived from theory.
\end{abstract}

@@

There do, however, also exist algorithms that allow us to parallelize simulations in the time domain. This could be very useful for running smaller simulations through a very long timespan. This is of interest for example in solar system dynamics. If we want to learn more about the behavior of mostly stable planetary orbits, we have to simulate a long time. Thus, parallel-in-time algorithms can be employed.\cite{parallelsolar}

In particle simulations, we are often numerically solving an ODE in the form:
$$
\diff{u(t)}{t} = f(t, u(t))
$$
where $f$ is the function describing our problem, $u(t)$ is the state of the system at time $t$ and we are given an initial condition $u(t_0) = u_0$.

@@

The parareal algorithm was introduced by \cite{parareal}.\footnote{The original paper in French, however it includes a description in English.} It achieves parallel-in-time integration using two types of integrators. A coarse solver $\mathcal{C}$ is used first to roughly estimate the solution, while a fine (but potentially expensive) solver $\mathcal{F}$ is then used to fill in the gaps in this estimated trajectory.

@@

\subsection*{Stopping criteria}
We assume the algorithm has converged enough when $\forall n: |u^{i+1}_n - u^i_n| < \varepsilon$. Other criteria also exist.\cite{speedup}

@@

Thus, to reach a high speed-up, we want the ratio $\frac{T_\mathcal{C}}{T_\mathcal{F}}$ to be small while also keeping the number of required iterations $k$ small.\cite{speedup}

The algorithm always converges in at most $N+1$ iterations, however, that is not useful (and takes even more time), because we have to converge at a significantly smaller number of iterations to gain any real speed-up.\cite{speedup}

@@

Thus, we are integrating the following equation:
$$
\diff{\mathbf{u}(t)}{t} = f(t, \mathbf{u}(t)) = f(t, (\mathbf{x}(t), \mathbf{v}(t))) = (\mathbf{v}(t), \mathbf{a}(t))
$$
where 
$$
a_i(t) = -G \sum_j m_j\frac{x_i(t)-x_j(t)}{\|x_i(t)-x_j(t)\|_2^3}
$$
where $m_j$ is the mass of the $j$-th particle. Acceleration is derived solely from the Newton's law of universal gravitation.

It is also possible to use a similar algorithm to the parareal algorithm using a simplified model instead of the coarse solver. This was done prior to the invention of parareal by \cite{parallelsolar} also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets.\cite{parallelsolar}\cite{farfuture}

@@

\subsection{Implementation}
An implementation in C++ using OpenMP for the parallelization was written. In case we would want to use a substantially higher number of CPUs, we would have to use a different framework, such as MPI. However, due to the higher complexity and difficulty of such an implementation, an easier approach using OpenMP was used in this project for the present time instead.

@@

@@

The resulting run times were collected in Table \ref{runtimes}. For visualizations of the simulation results see Figures \ref{sim} and \ref{energy}.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
\caption{Conservation of energy in the simulation. Same simulation as in Fig. \ref{sim}.}
\label{energy}
\end{figure}

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_5.eps}}
\caption{Comparison of the expected run times (blue curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red point)) and the real measured times with different numbers of CPUs (blue points).}
\label{speeds}
\end{figure}

\begin{table*}[htbp]
\caption{Comparison of running times taken by different configurations of the simulator. 
$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively. 
To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).
The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
\begin{center}
\begin{tabular}{crrrrrrrrrcc}
\toprule
Program & Time & n CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & $\Delta t_{\text{Final}}$ & Iters $^{\mathrm{b}}$ & Abs. err. $^{\mathrm{c}}$ & Rel. err. $^{\mathrm{d}}$ & CPU time & Real time \\
\midrule
%\multicolumn{11}{c}{Solar system $10^4\text{days}$} \\
%\midrule[.025em]
parareal & $10^4$ days & 64 & $0.5\times10^4$ &  $0.5\times10^5$ & 0.003125 day & 1 & 0.0110 AU & 3.60\% & 00:11:53 & 00:00:35 \\
parareal & $10^4$ days & 32 & $1\times10^4$ &  $1\times10^5$ & 0.003125 day & 1 & 0.0110 AU & 3.60\% & 00:12:53 & 00:00:52 \\
parareal & $10^4$ days & 16 & $2\times10^4$ &  $2\times10^5$ & 0.003125 day & 1 & 0.0109 AU & 3.60\% & 00:10:25 & 00:01:01 \\
parareal & $10^4$ days & 4 & $8\times10^4$ &  $8\times10^5$ & 0.003125 day & 1 & 0.0109 AU & 3.57\% & 00:10:27 & 00:02:54 \\
serial & $10^4$ days & 1 & N/A & $32\times10^5$ & 0.003125 day & N/A & 0 AU & 0\% & 00:05:01 & 00:05:01 \\
\bottomrule
\multicolumn{11}{l}{$^{\mathrm{a}}$ The number of segments ($N$) is the same as the number of CPUs.} \\
\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation}
\end{tabular}
\label{runtimes}
\end{center}
\end{table*}

@@

\subsection{Speed-up}
We measured the run times of a fixed simulation with different numbers of CPUs and compared them with the theoretical expectations. As was visualized in Figure \ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups). The implementation is consistently slower than the estimated run times, however, there are many factors that weren't considered in the speed-up estimation, for example we only took the function evaluations in the solvers into account. There are also other factors like synchronization costs of the OpenMP threads.

\subsection{Molecular dynamics}
We evaluated parareal only in the example of our solar system. Those settings, however, have their own specifics, in particular, planets in our solar system follow orbits with very different periods, which makes them harder to simulate and require very small step sizes.

This is not present in molecular dynamics settings and thus we could expect more stable behavior. This would be an interesting area to continue with this work.

We could further improve our simulations by further specializing them for solar system dynamics as was done in \cite{parallelsolar} (use a different coarse model simulating only the interactions of Sun and other planets). However, the direction of this paper is more about general particle simulations, so only the parareal integration was used.

@@

\section*{Acknowledgment}
\addcontentsline{toc}{section}{Acknowledgment}

@@

\begin{thebibliography}{00}
\bibitem{parareal}Lions, J., Maday, Y. \& Turinici, G. Résolution d'EDP par un schéma en temps «pararéel ». {\em Comptes Rendus De L'Académie Des Sciences - Series I - Mathematics}. \textbf{332}, 661-668 (2001), [Online]. Available: \url{https://www.sciencedirect.com/science/article/pii/S0764444200017936}
\bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
\bibitem{parallelsolar}Saha, P., Stadel, J. \& Tremaine, S. A Parallel Integration Method for Solar System Dynamics. {\em The Astronomical Journal}. \textbf{114} pp. 409 (1997,7), \url{http://dx.doi.org/10.1086/118485}
\bibitem{speedup}Pentland, K., Tamborrino, M., Sullivan, T., Buchanan, J. \& Appel, L. GParareal: a time-parallel ODE solver using Gaussian process emulation. {\em Statistics And Computing}. 33, 23 (2022,12), [Online]. Available: \url{https://doi.org/10.1007/s11222-022-10195-y}
\bibitem{farfuture}Gander, M. Is it possible to predict the far future before the near future is known accurately?. {\em Snapshots Of Modern Mathematics From Oberwolfach; 2019}. pp. 21 (2019), [Online]. Available: \url{http://publications.mfo.de/handle/mfo/3693}
\bibitem{parareal2}Gander, M. \& Vandewalle, S. ``Analysis of the Parareal Time‐Parallel Time‐Integration Method.'' SIAM Journal On Scientific Computing. 29, 556-578 (2007), [Online]. Available: \url{https://doi.org/10.1137/05064607X}
\end{thebibliography}

>>>>

\begin{abstract}
The purpose of this paper is to examine the applications of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration. We then test and measure the real speed-up and compare it with the expectations derived from theory.
\end{abstract}

@@

There do, however, also exist algorithms that allow us to parallelize simulations in the time domain. This could be very useful for running smaller simulations through a very long timespan. This is of interest for example in solar system dynamics.\footnote{Also in other fields such as molecular dynamics \cite{moldyn} but solar system simulations are a relatively good example.} If we want to learn more about the behavior of mostly stable planetary orbits, we have to simulate a long time. Thus, parallel-in-time algorithms can be employed. \cite{parallelsolar}

In particle simulations, we are often under the hood numerically solving an ODE in the form:
$$
\diff{u(t)}{t} = f(t, u(t))
$$
where $f$ is the function describing our problem, $u(t)$ is the state of the system at time $t$ and we are given an initial condition $u(t_0) = u_0$.

@@

The parareal algorithm was introduced by Lions, Maday and Turinici in 2001. \cite{parareal}\footnote{The original paper is in French but it does include a basic description in English.} It achieves parallel-in-time integration using two types of integrators. A coarse solver $\mathcal{C}$ is used first to roughly estimate the solution, while a fine (but potentially expensive) solver $\mathcal{F}$ is then used to fill in the gaps in this estimated trajectory.

@@

\subsection*{Stopping criteria}
We assume the algorithm has converged enough when $\forall n\colon |u^{i+1}_n - u^i_n| < \varepsilon$. Other criteria also exist. \cite{speedup}

@@

Thus, to reach a high speed-up, we want the ratio $\frac{T_\mathcal{C}}{T_\mathcal{F}}$ to be small while also keeping the number of required iterations $k$ small. \cite{speedup}

The algorithm always converges in at most $N+1$ iterations, however, that is not useful (and takes even more time), because we have to converge at a significantly smaller number of iterations to gain any real speed-up. \cite{speedup}

@@

Thus, we are integrating the following equation:
$$
\diff{\mathbf{u}(t)}{t} = f(t, \mathbf{u}(t)) = f(t, (\mathbf{x}(t), \mathbf{v}(t))) = (\mathbf{v}(t), \mathbf{a}(t))
$$
where 
$$
a_i(t) = -G \sum_j m_j\frac{x_i(t)-x_j(t)}{\|x_i(t)-x_j(t)\|_2^3}
$$
where $m_j$ is the mass of the $j$-th particle. Acceleration is derived solely from Newton's law of universal gravitation.

It is also possible to use a similar algorithm to the parareal algorithm using a simplified model instead of the coarse solver. This was done prior to the invention of parareal by \cite{parallelsolar} also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets. \cite{parallelsolar} \cite{farfuture}

@@

\subsection{Implementation}
An implementation in C++ using OpenMP for the parallelization was written.\footnote{The code is available at \url{https://github.com/lesves/mol-dyn}.} In case we would want to use a substantially higher number of CPUs, we would have to use a different framework, such as MPI. However, due to the higher complexity and difficulty of such  implementation, an simpler approach with OpenMP was used in this project instead.

@@

The experimental runs were conducted on the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization. Each computer in the cluster is equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.

@@

The resulting run times were collected in Table \ref{runtimes}. For visualizations of the simulation results see Figure \ref{sim} and \ref{energy}.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
\caption{Conservation of energy in the simulation. Same simulation as in Figure \ref{sim}. With a small enough step size, even a non-symplectic integrator like RK4 can preserve energy for some number of timesteps.}
\label{energy}
\end{figure}

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_5.eps}}
\caption{Comparison of the expected run times (blue curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red point)) and the real measured times with different numbers of CPUs (blue points). For more information, see Table \ref{runtimes}.}
\label{speeds}
\end{figure}

\begin{table*}[htbp]
\caption{Comparison of running times taken by different configurations of the simulator. 
$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively and $\Delta t_{\text{Final}}$ is the resulting step size.
The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
\begin{center}
\begin{tabular}{crrrrrrrrrcc}
\toprule
Program & Time & $n$ CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & $\Delta t_{\text{Final}}$ & Iters $^{\mathrm{b}}$ & Abs. err. $^{\mathrm{c}}$ & Rel. err. $^{\mathrm{d}}$ & CPU time & Real time \\
\midrule
%\multicolumn{11}{c}{Solar system $10^4\text{days}$} \\
%\midrule[.025em]
parareal & $10^4$ days & 64 & $0.5\times10^4$ &  $0.5\times10^5$ & 0.003125 day & 1 & 0.0110 AU & 3.60\% & 00:11:53 & 00:00:35 \\
parareal & $10^4$ days & 32 & $1\times10^4$ &  $1\times10^5$ & 0.003125 day & 1 & 0.0110 AU & 3.60\% & 00:12:53 & 00:00:52 \\
parareal & $10^4$ days & 16 & $2\times10^4$ &  $2\times10^5$ & 0.003125 day & 1 & 0.0109 AU & 3.60\% & 00:10:25 & 00:01:01 \\
parareal & $10^4$ days & 4 & $8\times10^4$ &  $8\times10^5$ & 0.003125 day & 1 & 0.0109 AU & 3.57\% & 00:10:27 & 00:02:54 \\
serial & $10^4$ days & 1 & N/A & $32\times10^5$ & 0.003125 day & N/A & 0 AU & 0\% & 00:05:01 & 00:05:01 \\
\bottomrule
\multicolumn{11}{l}{$^{\mathrm{a}}$ The number of segments ($N$) is the same as the number of CPUs.} \\
\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation}
\end{tabular}
\label{runtimes}
\end{center}
\end{table*}

@@

\subsection{Speed-up}
We measured the run times of a fixed simulation with different numbers of CPUs and compared them with the theoretical expectations. As was visualized in Figure \ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups).

The implementation is consistently slower than the estimated run times, however, there are many factors that weren't considered in the speed-up estimation, for example we only took the function evaluations in the solvers into account. There are also other factors such as the synchronization costs of the OpenMP threads.

In conclusion, we were able to achieve noticable speed-ups consistent with the theoretical foundation by using the parareal algorithm.

\subsection{Other particle simulations: Molecular dynamics}
We evaluated parareal only in the example of our solar system. Those settings, however, have their own specifics, in particular, planets in our solar system follow orbits with very different periods, which makes them harder to simulate and require very small step sizes.

This is usually not present in molecular dynamics settings and thus we could expect a more stable behavior. This could be an interesting area to continue with this work.

On the other hand, we could further improve our solar system simulations by further specializing them as was done in \cite{parallelsolar} (a different coarse model simulating only the interactions of Sun and other planets was used instead of a course integrator). However, our intent is to study general particle simulations, so this would be beyond our scope.

@@

\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

@@

\begin{thebibliography}{00}
\bibitem{parareal}Lions, J., Maday, Y. \& Turinici, G. Résolution d'EDP par un schéma en temps «pararéel ». {\em Comptes Rendus De L'Académie Des Sciences - Series I - Mathematics}. \textbf{332}, 661-668 (2001), [Online]. Available: \url{https://www.sciencedirect.com/science/article/pii/S0764444200017936}
\bibitem{parallelsolar}Saha, P., Stadel, J. \& Tremaine, S. A Parallel Integration Method for Solar System Dynamics. {\em The Astronomical Journal}. \textbf{114} pp. 409 (1997,7), \url{http://dx.doi.org/10.1086/118485}
\bibitem{speedup}Pentland, K., Tamborrino, M., Sullivan, T., Buchanan, J. \& Appel, L. GParareal: a time-parallel ODE solver using Gaussian process emulation. {\em Statistics And Computing}. 33, 23 (2022,12), [Online]. Available: \url{https://doi.org/10.1007/s11222-022-10195-y}
\bibitem{parareal2}Gander, M. \& Vandewalle, S. ``Analysis of the Parareal Time‐Parallel Time‐Integration Method.'' SIAM Journal On Scientific Computing. 29, 556-578 (2007), [Online]. Available: \url{https://doi.org/10.1137/05064607X}
\bibitem{farfuture}Gander, M. Is it possible to predict the far future before the near future is known accurately?. {\em Snapshots Of Modern Mathematics From Oberwolfach; 2019}. pp. 21 (2019), [Online]. Available: \url{http://publications.mfo.de/handle/mfo/3693}
\bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
\bibitem{moldyn}Baffico, L., Bernard, S., Maday, Y., Turinici, G. \& Zérah, G. Parallel-in-time molecular-dynamics simulations. {\em Physical Review E}. \textbf{66} (2002,11), http://dx.doi.org/10.1103/physreve.66.057701

\end{thebibliography}
